Designing Types for R, Empirically

### 2020-05-15 paper/main.tex
Include macros file
6.2 edits

In the last section, we evaluated our type annotation language based on a
set of function types that were automatically generated by \typetracer.
Now, we're interested in knowing how representative our trace-generated
types are of the functions they were generated for.  To measure this, we
conducted a large-scale experiment: for each package in the corpus discussed
in Section~\ref{sec:corpus}, we ran \typetracer on its test, example, and
vignette code, obtaining types for package functions, and then subsequently
ran the test, example, and vignette code of each of the package's reverse
dependencies (or clients).  We collected failed contract assertions in order
to analyze them and perhaps identify weaknesses in our type annotation
language.

We ran our evaluation on \NUMPKGSEVAL packages and recorded \NUMASSERTIONS
total assertions.  Overall, we found that only \PERCFAILEDASSERTIONSWUNDEF
of contract assertions failed.  We made a simplification in our analysis,
capping the number of function arguments we generate types for to 20, and
this resulted in only \PERCASSERTIONSUNDEF failed assertions, which suggests
that the simplification we made did not greatly affect the results.  Once we
controlled for these, we found that \PERCFAILEDASSERTIONS of all assertions
failed for reasons related to our generated types.  Additionally, we found
that \PERCSUCCARG of parameter types and \PERCSUCCFUNS of functions types
were never violated.  The number of immaculate function types increases to
\PERCSUCCFUNNOSTHREE if we count out S3 functions: These S3 functions
represent user-defined polymorphism which we do not tackle in this work.
Overall, these numbers are promising, and suggest that the type signatures
generated by running \typetracer on our corpus are indeed representative,
and capture intended function behaviour.

We breakdown the failed assertions by type in
Table~\ref{tbl:fail-breakdown-0}.  By far the most commonly failed assertion
is one asserting that a value with ${\bf dbl}[]$ type has type ${\bf
  class}\langle matrix \rangle$.  This was a deliberate design decision, as
coercion from vectors to matrices is ad hoc at best, and surely not a
practice codified in the language.  In a similar vein, another popular
failing assertion is checking if a ${\bf dbl}[]$ has type ${\bf int}[]$,
another case of commonly performed coercion.  We did not include these types
of coercions in our type annotation framework as programmers cannot rely on
them, and it is not always the case that the coercions are safe to perform.

@@

Turning our attention now to arguments, we found that only
\PROPARGSFAILINGASSERTS of function parameters exhibited a failed assertion.
Table~\ref{tbl:fail-breakdown-0} showed the raw results from counting
runtime occurrences, but that data alone does not tell the full story, as
some failures may be overrepresented if e.g. a failing contract assertion
was in a loop.  We were interested in knowing for each of the most common
violations in Table~\ref{tbl:fail-breakdown-0}, how many different arguments
had that type, and how many of those exhibited the contract failure in
question.  Table~\ref{tbl:fail-breakdown-1} breaks down the results from
Table~\ref{tbl:fail-breakdown-0} by unique parameter position, folding away
multiple identical failed contract assertions for the same parameter
position.  The first row of this table reads: a value of type
\code{double[]} was passed to 15 different function parameters expecting a
matrix, of which there are 1394 in total: 1.08\% of matrix-typed parameters
were passed double vector values.

@@

Table~\ref{tbl:fail-breakdown-2} presents data on the most frequently
violated contracts amongst the most frequently occurring argument types.  We
selected argument types which were in the 90th percentile of argument type
occurrences, computed the most frequent type signature violations among
them, and reported the most frequently violated contracts together with the
type of the value that violated that contract.  The first row of the table
reads: 44 function arguments with integer vector type are passed double
vectors instead, and 657 arguments have integer vector type.  Had we failed
to capture some key usage pattern of R with our type annotation framework,
we would likely see it here, and we can see this in action if we consider
Table~\ref{tbl:fail-breakdown-3}, which was obtained identically to
Table~\ref{tbl:fail-breakdown-2} with a percentile of 80th percentile
instead.  The most frequent argument type violation pattern in that of data
frames being passed to arguments expecting tibbles: this occurs in only 22\%
of such arguments, and represents cases where tests did not adequately cover
all valid function inputs.  Separate from the issue of testing, we can
capture this behaviour with user-defined subtyping, as data frames are
subtypes of tibbles, but this necessitates a proper object-oriented type
system for R which is outside the scope of this work.

>>>>

In the last section, we evaluated our type annotation language based on a
set of function types that were automatically generated by \typetracer.
Now, we're interested in knowing how representative our trace-generated
types are of the functions they were generated for.  To measure this, we
conducted a large-scale experiment: for each package in the corpus discussed
in Section~\ref{sec:corpus}, we ran \typetracer on its test, example, and
vignette code, obtaining types for package functions. Then, we
ran the test, example, and vignette code of each of the package's reverse
dependencies (or clients).  We collected failed contract assertions in order
to analyze them and perhaps identify weaknesses in our type annotation
language.

We ran our evaluation on \NUMPKGSEVAL packages and recorded \NUMASSERTIONS
total assertions.  Overall, we found that only \PERCFAILEDASSERTIONSWUNDEF
of contract assertions failed.  We made a simplification in our analysis,
capping the number of function arguments we generate types for to 20, and
this resulted in only \PERCASSERTIONSUNDEF failed assertions, which suggests
that the simplification we made did not greatly affect the results.  Once we
controlled for these, we found that \PERCFAILEDASSERTIONS of all assertions
failed for reasons related to our generated types.  Additionally, we found
that \PERCSUCCARG of parameter types and \PERCSUCCFUNS of functions types
were never violated.  The number of immaculate function types increases to
\PERCSUCCFUNNOSTHREE if we discount S3 functions: These S3 functions
represent user-defined polymorphism which we do not tackle in this work.
Overall, these numbers are promising, and suggest that the type signatures
generated by running \typetracer on our corpus are indeed representative,
and capture intended function behaviour.

We breakdown the failed assertions by type in
Table~\ref{tbl:fail-breakdown-0}.  
Accounting for 27.37\% of assertion failures, the most common vilation arises when a value with \CLASS{matrix} is passed where a \VEC\dbl is expected.
Considering these types, we might imagine them to be compatible, however not allowing this coercion was a deliberate design decision:
coercion from vectors to matrices is ad hoc at best, and surely not a
practice codified in the language.  In a similar vein, another popular
failing assertion is checking if a ${\bf dbl}[]$ has type ${\bf int}[]$,
another case of commonly performed coercion.  We did not include these types
of coercions in our type annotation framework as programmers cannot rely on
them, and it is not always the case that the coercions are safe to perform.

@@

Turning our attention now to arguments, we found that only
\PROPARGSFAILINGASSERTS of function parameters exhibited a failed assertion.
Table~\ref{tbl:fail-breakdown-0} showed the raw results from counting
runtime occurrences, but that data alone does not tell the full story, as
some failures may be overrepresented if e.g. a failing contract assertion
was in a loop.  We were interested in knowing for each of the most common
violations in Table~\ref{tbl:fail-breakdown-0}, how many different arguments
had that type, and how many of those exhibited the contract failure in
question.  Table~\ref{tbl:fail-breakdown-1} breaks down the failed assertions by type,
folding away multiple identical failed contract assertions for the same parameter
position.  The first row of this table reads: a value of type
\code{double[]} was passed to 15 different function parameters expecting a
matrix, of which there are 1394 in total: 1.08\% of matrix-typed parameters
were passed double vector values.

@@

Table~\ref{tbl:fail-breakdown-2} presents data on the most frequently
violated contracts amongst the most frequently occurring argument types.  We
selected argument types which were in the 90th percentile of argument type
occurrences, computed the most frequent type signature violations among
them, and reported the most frequently violated contracts together with the
type of the value that violated that contract.  The first row of the table
reads: 44 function arguments with \VEC\int type are passed \VEC\dbl  instead, and 657 arguments have \VEC\int type.  Had we failed
to capture some key usage pattern of R with our type annotation framework,
we would likely see it here, and we can see this in action if we consider
Table~\ref{tbl:fail-breakdown-3}, which was obtained identically to
Table~\ref{tbl:fail-breakdown-2} with a percentile of 80th percentile
instead.  The most frequent argument type violation pattern in that of data
frames being passed to arguments expecting tibbles: this occurs in only 22\%
of such arguments, and represents cases where tests did not adequately cover
all valid function inputs.  Separate from the issue of testing, we can
capture this behaviour with user-defined subtyping, as data frames are
subtypes of tibbles, but this necessitates a proper object-oriented type
system for R which is outside the scope of this work.
